Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-04 aistats/paper.tex
attempt at a motivation

% my packages
\usepackage{../math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\natp}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size
\newcommand{\lin}[1]{\left\langle#1\right\rangle}

@@

\section{Introduction and Background}

\fdk{Consider the optimization of 
\begin{equation}
	f(\theta) = \expect[x \sim p(x)]{\frac{1}{2}\norm{\theta}^2 - \lin{x, \theta}},
\end{equation}
given samples from $x$. We know how to do that one, 
and the worst-case scenario happens when $p(x)$ is a Gaussian.
Now how about the slightly more general case where
\begin{equation}
	f(\theta) = \expect[x \sim p(x)]{A(\theta) - \lin{x, \theta}}
\end{equation}
and we have full knowledge of $A$, a strictly convex twice differentiable function. 
How hard can that be?
Turns out, if $p$ is the maximum entropy distribution with 
$\expect[x \sim p(x)]{x} = \mu$, 
then it's a bit hard.}

Call both to stats and optimization crowds by introducing the parallel quite early.

@@

\begin{itemize}
	\item Stochastic and online mirror descent: 
	We obviously have rates in this setting, but they all assume 
	that the reference function is strongly-convex relative to a norm. 
	\citep[e.g.][Thm 4.2]{bubeck2015convex}
	\item What is mirror descent called in the literature?
		\begin{itemize}
			\item Mirror Descent \citep{nemirovski1983problem,beck2003mirror} 
			\item NoLips \citep{bauschke2017descent}
		\end{itemize}
	\item Interest in mirror descent boosted by the relatively smooth view 
	of \citet{bauschke2017descent,lu2018relatively}	
	\item 
	The oroblems on how to deal with Bregman divergences abound in optimization, 
	beyond stochasticity. 
	For example, we haven't yet figured out the analog of Nesterov-type acceleration 
	on relatively smooth and strongly convex problems, 
	to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$, 
	or just from $1/T$ to $1/T^2$ 
	in the (non-strongly) convex case.
	\citet{dragomir2021optimal} show that naïve application of Bregman updates can not achieve acceleration.
	The tools developped to make progress on one problem might help make progress on the other. 
	\item 
	Stochastic mirror descent in the relatively smooth setting;
	\citet{hanzely2018fastest},
	\citet{dragomir2021fast}
	and the zoo of assumptions
	\item Something about the statistical asymptotic behavior for the MLE? 
	\begin{equation}
		D_A(\theta_*, \theta_n) \approx \frac{1}{2}\norm{\theta_n - \theta_*}_{\nabla^2 A(\theta_*)}^2 
		\approx \frac{1}{n}?
	\end{equation}
	\item Something about how the Bregman divergence is often not considered in the statistics lit 
	because it can be infinite and is non-symmetric and not nice to work with, 
	and so folks focus on $\expect{\norm{\theta_n - \theta_*}^2}$ even if it's somewhat meaningless 
	in the non-asymptotic regime, 
	and the best non-asymptotic results 
	still rely on Taylor expansions after $n > $ something big?
	\item Something about how we don't even know how to deal with decreasing step-sizes 
	of the order of $1/t$ 
	\item 
	Should cite Bach's line of work on SGD, averaging schemes and self-concordant functions https://www.di.ens.fr/~fbach/bach_ejs_self_concordance.pdf
	that obtain the $1/t$ rate in the smooth, convex setting instead of $1/\sqrt{t}$, 
	which is the ``expected'' rate for that setting, 
	and how this approach above is a generalization beyond the Euclidean noise model 
	http://proceedings.mlr.press/v99/marteau-ferey19a/marteau-ferey19a.pdf
	http://proceedings.mlr.press/v99/ostrovskii19a/ostrovskii19a.pdf
\end{itemize}

>>>>

% my packages
\usepackage{../math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\cond}{\,\vert\,}
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\natp}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size
\newcommand{\lin}[1]{\left\langle#1\right\rangle}

@@

\section{Motivation}

We kind of know how to do optimization 
on smooth, strongly convex functions 
when there is some bound on the stochasticity,
for example in the form of a bounded variance, 
say for the problem 
\begin{equation}
\min_\theta f(\theta)
\end{equation}
with access to stochastic gradients 
$g(\theta)$ such that 
\begin{equation}
	\expect{\norm{g(\theta) - \nabla f(\theta)}^2} \leq \sigma^2.
\end{equation}
The best convergence rate, of stochastic gradient methods 
is obtained by SGD with decreasing step-sizes and averaging, 
yields a convergence rate of $1/t$ and is asymptotically tight;
SGD achieves this rate on the following problem, 
\begin{equation}
	f(\theta) = \frac{1}{2}\norm{\theta -\mu}^2, \quad
	g(\theta) \sim \sim \cN(\theta - \mu, \sigma^2).
\end{equation}
This problem is estimating the mean of a Gaussian, 
and it is not improvable by information theoretic arguments.

But our understanding of stochastic optimization lacks in other domains. 
A generalization of the above problem, 
at the intersection of optimization and estimation, 
is the problem of minimizing 
\begin{equation}
	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{A(\theta) - \lin{t(x), \theta}},
\end{equation}
where $A$ is a convex, Legendre function and 
$p(x\cond\theta)$ is the exponential family distribution 
with $A$ as its log-partition function. 
The Gaussian example above is a special case 
with $A(\theta) = \frac{1}{2}\norm{\theta - \mu}^2$
and $p(x\cond\theta) = \cN{\mu, \sigma^2}$.

We obviously have algorithms for this setting, 
as this is an maximum likelihood problem.
Giving the (online) maximum likelihood after seeing $t$ samples
which corresponds as online mirror descent with a decreasing step-size schedule (more on this later).
But our convergence rates are weak.
There are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.

We do not have a solution. 
The goal of this paper, and the reason it is submitted at AISTATS, 
is to bring this issue to the attention of 
the community working at the intersection of machine learning, optimization and statistics community,
in the hope that we can make progress on this question;
Can we build algorithms for stochastic estimation 
when the objective function is not (close to) quadratic
and the noise is not (sub-)Gaussian?

We give motivating examples where solution of this problem would enable progress in other areas,
principally motivated by stochastic optimization, 
show the connections between the optimization and statistical estimation sides of the problem,
and discuss where our current theory fails to fully capture the problem.

We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate---this is part of the point of this paper---and that the statistical theory results for the convergence of the MLE to the solution 
are well known. 
Our point is more subtle, in that for the purposes of optimization, 
to mimic the results that are known for SGD in Euclidean settings, 
we would like a result of the form 
\begin{equation*}
	\text{optimality gap after $t$ steps}
	\leq 
	\frac{\text{``bias''}}{t^2}
	+ \frac{\text{``variance''}}{t},
\end{equation*}
which is the bit that is lacking.

Then; 
- description of the setting
- statistical solutions (asymptotic, l1 or l2 error)
- optimization solutions (online+bounded gradient, self-concordance, more recent relative smoothness results)
- potential application of ideas that would solve this, beyond this problem (acceleration, causality?)
- conclusion?

\section{Technical background}

@@

\begin{itemize}
	\item Stochastic and online mirror descent: 
	We obviously have rates in this setting, but they all assume 
	that the reference function is strongly-convex relative to a norm. 
	\citep[e.g.][Thm 4.2]{bubeck2015convex}
	\item What is mirror descent called in the literature?
		\begin{itemize}
			\item Mirror Descent \citep{nemirovski1983problem,beck2003mirror} 
			\item NoLips \citep{bauschke2017descent}
		\end{itemize}
	\item Interest in mirror descent boosted by the relatively smooth view 
	of \citet{bauschke2017descent,lu2018relatively}	
	\item 
	The oroblems on how to deal with Bregman divergences abound in optimization, 
	beyond stochasticity. 
	For example, we haven't yet figured out the analog of Nesterov-type acceleration 
	on relatively smooth and strongly convex problems, 
	to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$, 
	or just from $1/T$ to $1/T^2$ 
	in the (non-strongly) convex case.
	\citet{dragomir2021optimal} show that naïve application of Bregman updates can not achieve acceleration.
	The tools developped to make progress on one problem might help make progress on the other. 
	\item 
	Stochastic mirror descent in the relatively smooth setting;
	\citet{hanzely2018fastest},
	\citet{dragomir2021fast}
	and the zoo of assumptions
	\item Something about the statistical asymptotic behavior for the MLE? 
	\begin{equation}
		D_A(\theta_*, \theta_n) \approx \frac{1}{2}\norm{\theta_n - \theta_*}_{\nabla^2 A(\theta_*)}^2 
		\approx \frac{1}{n}?
	\end{equation}
	\item Something about how the Bregman divergence is often not considered in the statistics lit 
	because it can be infinite and is non-symmetric and not nice to work with, 
	and so folks focus on $\expect{\norm{\theta_n - \theta_*}^2}$ even if it's somewhat meaningless 
	in the non-asymptotic regime, 
	and the best non-asymptotic results 
	still rely on Taylor expansions after $n > $ something big?
	\item Something about how we don't even know how to deal with decreasing step-sizes 
	of the order of $1/t$ 
	\item 
	Should cite Bach's line of work on SGD, averaging schemes and self-concordant functions 
	\url{https://www.di.ens.fr/~fbach/bach_ejs_self_concordance.pdf}
	that obtain the $1/t$ rate in the smooth, convex setting instead of $1/\sqrt{t}$, 
	which is the ``expected'' rate for that setting, 
	and how this approach above is a generalization beyond the Euclidean noise model 
	\url{http://proceedings.mlr.press/v99/marteau-ferey19a/marteau-ferey19a.pdf}
	\url{http://proceedings.mlr.press/v99/ostrovskii19a/ostrovskii19a.pdf}
\end{itemize}
